Master's Project Journal

### 2019-04-24 thesis/Background.tex
done done

Alternative designs replace the electrical lines used in the tree networks with waveguides for optical signals, with only the distribution in the local area carried out using regular wires. This technique presents many advantages \cite{chen2006chip}: optical clock delivery is immune to the noise sources that affect electrical clock distribution systems, consume less power and do not suffer from the electrical losses present in a regular tree system. However, this is a fledgling method, and sufficient research into materials and potential designs has yet to be carried out \ref{zianbetov2013phd,1648640}.

>>>>

Alternative designs replace the electrical lines used in the tree networks with waveguides for optical signals, with only the distribution in the local area carried out using regular wires. This technique presents many advantages: optical clock delivery is immune to the noise sources that affect electrical clock distribution systems, consume less power and do not suffer from the electrical losses present in a regular tree system \cite{chen2006chip}. However, this is a fledgling method, and sufficient research into materials and potential designs has yet to be carried out \cite{zianbetov2013phd,chen2006chip}.

### 2019-04-24 thesis/FPGABasedADPLLs.tex
done done

\section{Chapter Overview}
The first step in creating \iac{FPGA} based network of \acfp{ADPLL} is the design of the \ac{ADPLL} itself, which will be addressed in this chapter. The nature of \iac{FPGA} necessitates compromises in the design of a certain blocks, limiting direct transfer from \ac{ASIC} designs. In this chapter the potential designs for each individual block, or module, investigated will be explained and the case for their selection in \iac{FPGA} based \ac{ADPLL} made. A number of blocks have implemented by purely digital circuitry and as such can be transferred in their entirety from \iac{FPGA} and vice versa. However, those that will be used to emulate mixed-signal circuitry, such as the \ac{DCO} and \ac{PFD}, will be examined in greater detail. \Iac{RTL} diagram of the \ac{ADPLL} topology that will be used in the following sections is shown in Figure \ref{fig:network_adpll}.
\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{../network_adpll}
	\caption{\acs{ADPLL} designed for use in a Cartesian grid network.}
	\label{fig:network_adpll}
\end{figure}

@@

\subsection{\acs{FPGA} Driven, Linear Period \acs{DCO}}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../osc1}
	\caption{\acs{FPGA} driven, linear period \acs{DCO} \ac{RTL} Diagram.}
	\label{fig:osc1}
\end{figure}
The first design of \ac{DCO} to be examined, is of the type used by Zianbetov in his \ac{ADPLL} network test bed and relies on a counter driven by the clock manager on the \ac{FPGA} \cite{zianbetov2013phd}. At each event on the \ac{FPGA} provided clock, a counter is incremented, overflowing upon increment past the maximum possible value. The \ac{MSB} of this counter forms the waveform generated by this oscillator, the period of which is controlled through an adjustable value that is loaded into the counter once overflow is reached, and forms the starting point for the counter. The period of oscillation is given by:
\begin{equation}
	T_{osc} = \big(2^{width} -~(BIAS+CC)~\big)\times T_{FPGA}
\end{equation}
where $T_{FPGA}$ is the clock period of the \ac{FPGA}, $CC$ is the control code input, $BIAS$ centres the oscillator in the middle of the tuning range in the event that the control code is zero and $width$ is the width of the counter used. As the only variable here is the control code, the period step of this design is $T_{FPGA}$, thereby providing period linearity with respect to the control code. This is the key advantage of this design, as most \ac{ASIC} implementations of \iac{DCO} are also linear in period. The other main reason to choose this design, is that its \ac{FPGA} clocked nature allows for exact control over the frequency of operation, and the number of tunable parameters make it possible to configure multiple ways to achieve the same frequencies of operation. Combined, these attributes make it very easy to create an oscillator that emulates the behaviour of a design intended for \iac{ASIC}, however, at a greatly reduced frequency. This restriction on the frequency of operation arises out of the period step size, which in order to obtain a good resolution must be orders of magnitude smaller than the intended period. As the output waveform is taken from the counter's \ac{MSB}, the reload value of the counter must never go beyond $2^{width-1}$, as otherwise the output waveform will become a constant 1. As the reload value varies the low time of the \ac{MSB}, if the desired output waveform is a square wave, this design will not be suitable.

@@

To determine the sign itself, the clock edge which first had a rising edge must be recorded, and this can be done by setting \iac{SR} latch if the rising edge of the generated signal occurs first and resetting in the case of the reference clock having the first rising edge. Looking at the whole measurement cycle, however, it is apparent that all four combinations possible of two inputs occur, putting the \ac{SR} latch into which the output is undefined. This problem can be avoided through careful design. Examining the truth table in Table \ref{table:sr_tt}, the combination to be avoided is the presence of a 1 on both outputs, precisely the situation seen at the end of the measurement interval. In order to avoid the undefined state occurring at the end of the measurement interval, both inputs and outputs to the \ac{SR} latch can be inverted using a NOT gate. Table \ref{table:sr_tt} also illustrates how the $\overline{Q}$ output now behaves as $Q$ did previously, however, the end of measurement state where both input registers have been triggered now causes the \ac{SR} latch to enter the hold state, where it maintains the previous values. The sign, $\overline{Q}$ of the \ac{SR}, can then be loaded into a register using the \texttt{clear} signal generated previously, to maintain it until the next measurement completes.
\begin{table}[ht]
	\begin{center}
		\begin{tabular}{cc|cc}           
			$S$&$R$&$Q$&$\overline{Q}$\\
			\hline
			0&0&$Q_{n-1}$&$\overline{Q}_{n-1}$\T\\
			0&1&0&1\T\\
			1&0&1&0\T\\
			1&1&\multicolumn{2}{c}{undefined}\T\\					
		\end{tabular}
		\caption{SR latch truth table}
		\label{table:sr_tt}
	\end{center}
\end{table}

This, however, ignores the potential for metastability which may arise if both signals have a rising edge within an extremely short period of time. While it may be feasible in other systems to dismiss this occurrence as a low probability event, the goal of \iac{PLL} is to reduce the time delay between rising edges to zero. This increases the possibility of a collision occurring, which may lead to indeterminate behaviour in other parts of the feedback network. In order to avoid this in \iac{ASIC}, an arbitration circuit is placed between the \ac{SR} latch and the register storing the sign between measurements, which will default to one of the two signals in this case. This is, however, a mixed signal circuit, such as that proposed by Tierno \textit{et al} (2008) and implemented by Zianbetov and Shan \cite{tierno2008wide,zianbetov2013phd,shan2014phd}, which requires two \ac{PMOS} and two \ac{NMOS} transistors and cannot be implemented on \iac{FPGA}. As the output of the \ac{SR} latch will be indeterminate prior to the detection of the first edge, the impact of both edge detection registers' output is that the \ac{SR} latch attempts to hold this indeterminate value as an output. To avoid this unwanted behaviour the arbiter present on \iac{ASIC} is replaced by a second \ac{SR} latch, connected as in Figure \ref{fig:arbitration}. This ensures the value presented to the register storing the sign is always determinate.
\begin{table}[!ht]
	\begin{center}
		\setlength{\tabcolsep}{.5\tabcolsep}
		\begin{tabular}{l|cc|cc|c|cc|c}           
			When&$Gen$&$Ref$&$\overline{Gen}/S^a$&$\overline{Ref}/R^a$&$Q^a$&$S^b$&$R^b$&$Sign/Q^b$\\
			\hline
			Between Measurements&0&0&1&1&\multicolumn{1}{c|}{undefined}&\multicolumn{2}{c|}{01 or 10}&\multicolumn{1}{c}{0 or 1}\T\\
			Ref. First&0&1&1&0&1&0&1&0\T\\
			Gen. First&1&0&0&1&0&1&0&1\T\\	
			Measurement Complete&1&1&0&0&$Q^a_{n-1}$&$\overline{Q}^a_{n-1}$&$Q^a_{n-1}$&$\overline{Q}^a_{n-1}$\T\\					
		\end{tabular}
		\caption{Sign detector truth table}
		\label{table:sign_tt}
	\end{center}
\end{table}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{../simple_sign_detection}
	\caption[Example Sign Detector]{Example Sign Detector.}
	\label{fig:arbitration}
\end{figure}

@@

Again, both integer and fixed-point methods of \ac{LF} design are possible, with the integer method requiring the logic combining the two paths to preserve the relationship between the proportional and integral gains. From the work of Koskin \textit{et al} this known to be several binary orders of magnitude \cite{koskin2018generation}. Failure to preserve a relationship of this order may lead to the \ac{ADPLL} becoming unstable, as in the upper-left region of Figure \ref{fig:gain_grid}. In the fixed-point system the relationship between \ac{kp} and \ac{ki} is preserved during the combination of proportional and integral paths, as both values are aligned using their radix point, before addition is carried out.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../eugene}
	\caption[Stability of \acl{LF} gains]{Stability of \acl{LF} gains \cite{koskin2018generation}.}
	\label{fig:gain_grid}
\end{figure}

>>>>

\section{Chapter Overview}
The first step in creating \iac{FPGA} based network of \acfp{ADPLL} is the design of the \ac{ADPLL} itself, which will be addressed in this chapter. The nature of \iac{FPGA} necessitates compromises in the design of a certain blocks, limiting direct transfer from \ac{ASIC} designs. In this chapter the potential designs for each individual block, or module, investigated will be explained and the case for their selection in \iac{FPGA} based \ac{ADPLL} made. A number of blocks have implemented by purely digital circuitry and as such can be transferred in their entirety from \iac{FPGA} and vice versa. However, those that will be used to emulate mixed-signal circuitry, such as the \ac{DCO} and \ac{PFD}, will be examined in greater detail. \Iacf{RTL} diagram of the \ac{ADPLL} topology that will be used in the following sections is shown in Figure \ref{fig:network_adpll}.
\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth]{../network_adpll}
	\caption{\acs{ADPLL} designed for use in a Cartesian grid network.}
	\label{fig:network_adpll}
\end{figure}

@@

\subsection{\acs{FPGA} Driven, Linear Period \acs{DCO}}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../osc1}
	\caption{\acs{FPGA} driven, linear period \acs{DCO} \acs{RTL} Diagram.}
	\label{fig:osc1}
\end{figure}
The first design of \ac{DCO} to be examined, is of the type used by Zianbetov in his \ac{ADPLL} network test bed and relies on a counter driven by the clock manager on the \ac{FPGA} \cite{zianbetov2013phd}. At each event on the \ac{FPGA} provided clock, a counter is incremented, overflowing upon increment past the maximum possible value. The \ac{MSB} of this counter forms the waveform generated by this oscillator, the period of which is controlled through an adjustable value that is loaded into the counter once overflow is reached, and forms the starting point for the counter. The period of oscillation is given by:
\begin{equation}
	T_{osc} = \big(2^{width} -~(BIAS+CC)~\big)\times T_{FPGA}
\end{equation}
where $T_{FPGA}$ is the clock period of the \ac{FPGA}, $CC$ is the control code input, $BIAS$ centres the oscillator in the middle of the tuning range in the event that the control code is zero and $width$ is the width of the counter used. As the only variable here is the control code, the period step of this design is $T_{FPGA}$, thereby providing period linearity with respect to the control code. This is the key advantage of this design, as most \ac{ASIC} implementations of \iac{DCO} are also linear in period. The other main reason to choose this design, is that its \ac{FPGA} clocked nature allows for exact control over the frequency of operation, and the number of tunable parameters make it possible to configure multiple ways to achieve the same frequencies of operation. Combined, these attributes make it very easy to create an oscillator that emulates the behaviour of a design intended for \iac{ASIC}, however, at a greatly reduced frequency. This restriction on the frequency of operation arises out of the period step size, which in order to obtain a good resolution must be orders of magnitude smaller than the intended period. As the output waveform is taken from the counter's \ac{MSB}, the reload value of the counter must never go beyond $2^{width-1}$, as otherwise the output waveform will become a constant 1. As the reload value varies the low time of the \ac{MSB}, if the desired output waveform is a square wave, this design will not be suitable.

@@

To determine the sign itself, the clock edge which first had a rising edge must be recorded, and this can be done by setting \iac{SR} latch if the rising edge of the generated signal occurs first and resetting in the case of the reference clock having the first rising edge. Looking at the whole measurement cycle, however, it is apparent that all four combinations possible of two inputs occur, putting the \ac{SR} latch into which the output is undefined. This problem can be avoided through careful design. Examining the truth table in Table \ref{table:sr_tt}, the combination to be avoided is the presence of a 1 on both outputs, precisely the situation seen at the end of the measurement interval. In order to avoid the undefined state occurring at the end of the measurement interval, both inputs and outputs to the \ac{SR} latch can be inverted using a NOT gate. Table \ref{table:sr_tt} also illustrates how the $\overline{Q}$ output now behaves as $Q$ did previously, however, the end of measurement state where both input registers have been triggered now causes the \ac{SR} latch to enter the hold state, where it maintains the previous values. The sign, $\overline{Q}$ of the \ac{SR}, can then be loaded into a register using the \texttt{clear} signal generated previously, to maintain it until the next measurement completes.
\begin{table}[ht]
	\begin{center}
		\begin{tabular}{cc|cc}           
			$S$&$R$&$Q$&$\overline{Q}$\\
			\hline
			0&0&$Q_{n-1}$&$\overline{Q}_{n-1}$\T\\
			0&1&0&1\T\\
			1&0&1&0\T\\
			1&1&\multicolumn{2}{c}{undefined}\T\\					
		\end{tabular}
		\caption[SR Latch Truth Table]{SR Latch Truth Table.}
		\label{table:sr_tt}
	\end{center}
\end{table}

This, however, ignores the potential for metastability which may arise if both signals have a rising edge within an extremely short period of time. While it may be feasible in other systems to dismiss this occurrence as a low probability event, the goal of \iac{PLL} is to reduce the time delay between rising edges to zero. This increases the possibility of a collision occurring, which may lead to indeterminate behaviour in other parts of the feedback network. In order to avoid this in \iac{ASIC}, an arbitration circuit is placed between the \ac{SR} latch and the register storing the sign between measurements, which will default to one of the two signals in this case. This is, however, a mixed signal circuit, such as that proposed by Tierno \textit{et al} (2008) and implemented by Zianbetov and Shan \cite{tierno2008wide,zianbetov2013phd,shan2014phd}, which requires two \ac{PMOS} and two \ac{NMOS} transistors and cannot be implemented on \iac{FPGA}. As the output of the \ac{SR} latch will be indeterminate prior to the detection of the first edge, the impact of both edge detection registers' output is that the \ac{SR} latch attempts to hold this indeterminate value as an output. To avoid this unwanted behaviour the arbiter present on \iac{ASIC} is replaced by a second \ac{SR} latch, connected as in Figure \ref{fig:arbitration}. This ensures the value presented to the register storing the sign is always determinate.
\begin{table}[!ht]
	\begin{center}
		\setlength{\tabcolsep}{.5\tabcolsep}
		\begin{tabular}{l|cc|cc|c|cc|c}           
			When&$Gen$&$Ref$&$\overline{Gen}/S^a$&$\overline{Ref}/R^a$&$Q^a$&$S^b$&$R^b$&$Sign/Q^b$\\
			\hline
			Between Measurements&0&0&1&1&\multicolumn{1}{c|}{undefined}&\multicolumn{2}{c|}{01 or 10}&\multicolumn{1}{c}{0 or 1}\T\\
			Ref. First&0&1&1&0&1&0&1&0\T\\
			Gen. First&1&0&0&1&0&1&0&1\T\\	
			Measurement Complete&1&1&0&0&$Q^a_{n-1}$&$\overline{Q}^a_{n-1}$&$Q^a_{n-1}$&$\overline{Q}^a_{n-1}$\T\\					
		\end{tabular}
		\caption[Sign Detector Truth Table]{Sign Detector Truth Table}
		\label{table:sign_tt}
	\end{center}
\end{table}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{../simple_sign_detection}
	\caption[Example Sign Detector]{Example Sign Detector.}
	\label{fig:arbitration}
\end{figure}

@@

Again, both integer and fixed-point methods of \ac{LF} design are possible, with the integer method requiring the logic combining the two paths to preserve the relationship between proportional and integral gains. From the work of Koskin \textit{et al} this known to be several binary orders of magnitude \cite{koskin2018generation}. Failure to preserve a relationship of this order may lead to the \ac{ADPLL} becoming unstable, as in the upper-left region of Figure \ref{fig:gain_grid}. In the fixed-point system the relationship between \ac{kp} and \ac{ki} is preserved during the combination of proportional and integral paths, as both values are aligned using their radix point, before addition is carried out.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../eugene}
	\caption[Oscillator jitter for various \acl{LF} gains]{Oscillator jitter for various \acl{LF} gains \cite{koskin2018generation}.}
	\label{fig:gain_grid}
\end{figure}
